---
title: "Homework #8: Boosting" 
author: "Richard 'Ricky' Kuehn"
format: ds6030hw-html
---

::: {style="background-color:yellow; color:red; display: block; border-color: black; padding:1em"}
This is an **independent assignment**. Do not discuss or work with classmates.
:::

```{r config, include=FALSE}
# Set global configurations and settings here
knitr::opts_chunk$set()                 # set global chunk options
ggplot2::theme_set(ggplot2::theme_bw()) # set ggplot2 theme
```

# Required R packages and Directories {.unnumbered .unlisted}

```{r packages, message=FALSE, warning=FALSE}
# data_url = "https://archive.ics.uci.edu/static/public/275/bike+sharing+dataset.zip"
library(tidyverse)
library(lubridate)
library(padr)
library(lightgbm)
```

# Problem 1: Bike Sharing Data

This homework will work with bike rental data from Washington D.C.

## a. Load data

Load the *hourly* `Bikesharing` data from the [UCI ML Repository](https://archive.ics.uci.edu/dataset/275/bike+sharing+dataset).

::: {.callout-note title="Solution"}
```{r}
hour <- read.csv("hour.csv")
hour <- hour |> mutate(dteday = as.Date(dteday))
head(hour)
```
:::

## b. Data Cleaning

Check out the variable descriptions in the [Additional Variable Information](https://archive.ics.uci.edu/dataset/275/bike+sharing+dataset). To prepare the data for modeling, do the following:

1.  Convert the `weathersit` to an *ordered factor*.
2.  Unnormalize `temp` and `atemp` and convert to Fahrenheit.
3.  Unnormalize `windspeed`.

::: {.callout-note title="Solution"}
**1.**

```{r}
table(hour$weathersit)
```

```{r}
hour$weathersit <- factor(hour$weathersit, levels = c(1, 2, 3, 4), 
                          labels = c("Clear", "Mist", "Light Precipitation", "Heavy Precipitation"),
                          ordered = TRUE)
```

```{r}
table(hour$weathersit)
```

**2.**

```{r}
summary(hour$temp)
summary(hour$atemp)
```

```{r}
# new_temp <- (og_temp - t_min)/(t_max - t_min)
# og_temp <- new_temp * (t_max - t_min) + t_min
t_min <- -8
t_max <- 39
hour$temp <- hour$temp * (t_max - t_min) + t_min

a.t_min <- -16
a.t_max <- 50
hour$atemp <- hour$atemp * (a.t_max - a.t_min) + a.t_min
```

```{r}
summary(hour$temp)
summary(hour$atemp)
```

**3.**

```{r}
summary(hour$windspeed)
```

```{r}
# The values are divided to 67 (max)
hour$windspeed <- hour$windspeed * 67
```

```{r}
summary(hour$windspeed)
```
:::

## c. Missing times

Not every hour of every day is represented in these data. Some times, like 2011-03-15 hr=3, is due to daylight savings time. Other times, like 2011-01-02 hr=5, is probably due to the data collection process which ignored any times when `cnt = 0`.

This may not be perfect, but do the following to account for missing times:

1.  Create new rows/observations for all missing date-hr combinations that we think are due to actual zero counts. That is, exclude daylight savings. Set the outcome variables to zero (`causal = 0`, `registered = 0`, and `cnt = 0`) for these new observations. `tidyr::complete()` can help.

2.  Fill in the other missing feature values with values from previous hour. For example, the `temp` for 2011-01-02 **hr=5** should be set to the `temp` from the non-missing 2011-01-02 **hr=4**. `tidyr::fill()` can help.

::: {.callout-note title="Solution"}
```{r}
table(hour$hr)
```

```{r}
# find missing hours
hour_analysis <- hour |>
  arrange(dteday, hr) |>
  group_by(dteday) |>
  mutate(
    prev_hr = lag(hr),
    hr_gap = hr - coalesce(prev_hr, hr - 1)
  ) |>
  ungroup()

missing_hours <- hour_analysis |>
  filter(hr_gap > 1) |>
  select(dteday, hr, prev_hr, hr_gap)
```

```{r}
missing_hours
```

```{r}
# exclude daylight savings, complete missing hours using values from below, replace na for last 3 columns with 0
hour2 <- hour %>%
  complete(
    dteday = unique(dteday),
    hr = 0:23
  ) %>%
  filter(!(
    (dteday == '2011-03-13' & hr == 2) |
    (dteday == '2012-03-11' & hr == 2)
  )) %>%
  fill(instant, season, yr, mnth, holiday, weekday, workingday, 
       weathersit, temp, atemp, hum, windspeed) %>%
  mutate(
    casual = replace_na(casual, 0),
    registered = replace_na(registered, 0),
    cnt = replace_na(cnt, 0)
  )
```

```{r}
# check to make sure same count of hour values minus 2 for hour 2 because of 'leap forward'
table(hour2$hr)
```

```{r}
# convert to df, update instant column, set row names
hour2 <- hour2 %>%
  arrange(dteday, hr) %>%
  mutate(instant = 1:n()) %>%
  as.data.frame()

row.names(hour2) <- hour2$instant
hour2 <- hour2 %>% select(-instant)
```

```{r}
head(hour2, n=72)
```
:::

## d. New predictors

1.  Add the variable `doy` to represent the day of the year (1-366).
2.  Add the variable `days` to represent the *fractional number of days* since `2011-01-01`. For example hr=2 of 2011-01-02 is `r round(1 + 2/24, 3)`.
3.  Add lagged counts: autoregressive. Add the variable `cnt_ar` to be the `cnt` in the previous hour. You will need to set the value for `cnt_ar` for the 1st observation.\
4.  Add lagged counts: same time previous day, or a lag of 24 hours. You will need to set the values for the first 24 hours.

Hints:

-   The `lubridate` package (part of `tidymodels`) is useful for dealing with dates and times.
-   `dplyr::lag()` can help with making the lagged variables.

::: {.callout-note title="Solution"}
```{r}
hour2 <- hour2 %>%
  mutate(
    doy = yday(dteday),
    days = round(as.numeric(difftime(dteday, as.Date("2011-01-01"), units = "days")) + hr/24, 3)
  ) %>%
  arrange(dteday, hr) %>%
  mutate(
    cnt_ar = lag(cnt),
    cnt_lag24 = lag(cnt, 24)
  )

hour2$cnt_ar[1] <- 0
hour2$cnt_lag24[1:24] <- 0
```

```{r}
head(hour2, n = 48)
```
:::

## e. Train-Test split

Randomly select 1000 observations for the test set and use the remaining for training.

::: {.callout-note title="Solution"}
```{r}
set.seed(123)
test_ind <- sample(1:nrow(hour2), 1000)
train <- hour2[-test_ind,]
test <- hour2[test_ind,]
```
:::

# Problem 2: Predicting bike rentals

## a. Poisson loss

The outcome variables, number of renters, are counts (i.e., non-negative integers). For count data, the variance often scales with the expected count. One way to accommodate this is to model the counts as a Poisson distribution with rate $\lambda_i = \lambda(x_i)$. In lightgbm, the "poisson" objective uses an ensemble of trees to model the *log of the rate* $F(x) = \log \lambda(x)$. The poisson loss function (negative log likelihood) for prediction $F_i = \log \lambda_i$ is $\ell(y_i, F_i) = -y_iF_i + e^{F_i}$ where $y_i$ is the count for observation $i$ and $F_i$ is the ensemble prediction.

-   Given the current prediction $\hat{F}_i$, what is the *gradient* and *hessian* for observation $i$?
-   Page 12 of the [Taylor Expansion notes](lectures/taylor-expansion.pdf) shows that each new iteration of boosting attempts to find the tree that minimizes $\sum_i w_i (z_i - \hat{f}(x_i))^2$. What are the values for $w_i$ and $z_i$ for the "poisson" objective (in terms of $\hat{\lambda}_i$ *or* $e^{\hat{F}_i}$).

::: {.callout-note title="Solution"}
gradient and hessian for Poisson loss are:

gradient: y_i - $\lambda_i$ = y_i - e\^F_i

hessian: $\lambda_i$ = e\^F_i

Taylor expansion minimization...

w_i = e\^F_i

z_i = (y_i - e\^F_i) / e\^F_i = y_i / e\^F_i - 1
:::

## b. LightGBM Tuning

Tune a lightgbm model on the training data to predict the number of total number of renters (`cnt`). Do *not* use `registered` or `causal` as predictors!

-   Use the "poisson" objective; this is a good starting place for count data. This sets the loss function to the negative Poisson log-likelihood.

-   You need to tune at least two parameters: one related to the complexity of the trees (e.g., tree depth) and another related to the complexity of the ensemble (e.g., number of trees/iterations). [LightGBM documentation on parameter tuning](https://lightgbm.readthedocs.io/en/latest/Parameters-Tuning.html). And [LightGBM list of all parameters](https://github.com/microsoft/LightGBM/blob/master/docs/Parameters.rst).

-   You are free to tune other parameters as well, just be cautious of how long you are willing to wait for results.

i.  List relevant tuning parameter values, even those left at their default values. Indicate which values are non-default (either through tuning or just selecting). You can get these from the `params` element of a fitted lightgbm model, e.g., `lgbm_fitted$params`.

ii. Indicate what method was used for tuning (e.g., type of cross-validation).

::: {.callout-note title="Solution"}
```{r}
feature_cols <- setdiff(names(train), 
                       c("registered", "casual", "cnt", "dteday"))
```
```{r}
train_matrix <- sapply(train[, feature_cols], as.numeric)
train_lgb <- lgb.Dataset(
  data = as.matrix(train_matrix),
  label = train$cnt,
  free_raw_data = FALSE
)
```
```{r}
params <- list(
  objective = "poisson",  # specified
  metric = "poisson",     # specified
  boosting = "gbdt",      # default
  num_leaves = 31,        # default          
  min_data_in_leaf = 20   # default
)
learning_rate <- c(0.01, 0.03, 0.5, 0.7)  
max_depths <- c(4, 6, 8, 10)           
num_iterations <- c(100, 250, 400, 550)
```
```{r}
cv_results <- expand.grid(
 learning_rate = learning_rate,
 max_depth = max_depths,
 nrounds = num_iterations,
 poisson_loss = NA
)
```
```{r}
for(i in 1:nrow(cv_results)) {
 current_params <- c(params,
                    list(learning_rate = cv_results$learning_rate[i],
                         max_depth = cv_results$max_depth[i]))
 
 cv <- lgb.cv(
   params = current_params,
   data = train_lgb,
   nfold = 5,
   nrounds = cv_results$nrounds[i],
   early_stopping_rounds = 10,
   verbose = -1
 )
 
 cv_results$poisson_loss[i] <- min(unlist(cv$record_evals$valid$poisson$eval))
}
```
```{r}
best_params <- cv_results[which.min(cv_results$poisson_loss), ]
best_params
```

```{r}
final_model <- lgb.train(
 params = c(params, 
           list(max_depth = best_params$max_depth,
                learning_rate = best_params$learning_rate)),
 data = train_lgb,
 nrounds = best_params$nrounds,
 verbose = -1
)
```

:::

## c. Evaluation

Make predictions on the test data and evaluate. Report the point estimate and 95% confidence interval for the poisson log loss *and* the mean absolute error.

::: {.callout-note title="Solution"}
```{r}
test_matrix <- sapply(test[, feature_cols], as.numeric)
predictions <- predict(final_model, as.matrix(test_matrix))
```
```{r}
poisson_loss <- function(y_true, y_pred) {
  -mean(y_true * log(y_pred) - y_pred)
}
pll <- poisson_loss(test$cnt, predictions)
```
```{r}
mae <- mean(abs(test$cnt - predictions))
pll_se <- sd(-test$cnt * log(predictions) + predictions)/sqrt(length(predictions))
mae_se <- sd(abs(test$cnt - predictions))/sqrt(length(predictions))
```
```{r}
cat(sprintf("Poisson Loss: %.3f [%.3f, %.3f]\n",
           pll, pll - 1.96*pll_se, pll + 1.96*pll_se))

cat(sprintf("MAE: %.1f [%.1f, %.1f]",
           mae, mae - 1.96*mae_se, mae + 1.96*mae_se))
```
:::



